---
title: "SQL-NoSQL analysis"
author: "Kathleen DeBrota"
date: "April 21, 2019"
output: html_document
---

This document is for some analysis and visualizations of the credit card data set for SQL/NoSQL final project. 

Guiding questions: What kind of person is most likely to default on their credit card payments? What demographic factor(s) are most important in predicting this? 

1. Are younger people more or less likely to default on payments?
2. Are higher-educated people more or less likely to default?
3. How is credit limit associated with likelihood to default? How is credit limit associated with age and education level?
4. At what bill value is someone more likely to default on their payment? Are there any demographic trends that emerge there?
5. Is the amount of one single bill more influential than whether or not the  person delayed their last bill in deciding whether someone will default?

```{r}
library(readxl)
library(dplyr)
library(lattice)
library(rcompanion)
library(ggplot2)
```

```{r}
setwd("C:/Users/kdebr/Documents/GitHub/sql-nosql")
credit<-read.csv(file='cleanData.csv')
head(credit)
```
Exploratory views of the data.
```{r}
summary(credit)
```
More of the respondents were women (18,112 vs 11,888 men). The majority were undergraduates (14,030) with some graduates (10,585) and fewer who were high-school educated, uneducated, unknown, or 'other'. The majority were not married: 15,964, but many were married: 13,659. A few were 'other' or unknown for marital status. The mean age of the respondents was 35.49 years, with the youngest participants at 21 years and oldest at 79 years. 
```{r}
#view thistograms of demographic data
hist(credit$AGE, main='Histogram of age distribution of respondents', ylab='Frequency', xlab='Age in years')

#Boxplot of the payment status for each of the 6 payments.
boxplot(credit$PAY_1, credit$PAY_2, credit$PAY_3, credit$PAY_4, credit$PAY_5, credit$PAY_6, main='Payment status of each of 6 payments', ylab='Status code', xlab='Payment number')

#Boxplot of payment amount for each of 6 payments.
boxplot(credit$BILL_AMT1, credit$BILL_AMT2, credit$BILL_AMT3, credit$BILL_AMT4, credit$BILL_AMT5, credit$BILL_AMT6, main='Bill amount of each of 6 payments', ylab='Bill amount (NT dollars)', xlab='Payment number')



```
Separate out just those who defaulted and who did not. Visualize general demographic differences
```{r}
will_default<-subset(credit, DEFAULT_PAYMENT_NEXT_MONTH==1)
z3countdef<-table(will_default$AGE, will_default$MARRIAGE)

will_pay<-subset(credit, DEFAULT_PAYMENT_NEXT_MONTH==0)
#countpay<- table(will_pay$AGE, will_pay$MARRIAGE)

credit$MARRIAGE<-as.factor(credit$MARRIAGE)
credit$DEFAULT_PAYMENT_NEXT_MONTH <-as.factor(credit$DEFAULT_PAYMENT_NEXT_MONTH)

par(mfrow=c(1,2))
#View ages of those who did and did not default.
hist(will_default$AGE, main='Age distribution: Will default', xlab='Age (years)', ylab='Frequency')
hist(will_pay$AGE, main='Age distribution: Will not default', xlab='Age(years)', ylab='Frequency')

#----------------------------------------------

#View marital status of those who did/didn't default.
par(mfrow=c(1,1))
marriage <-credit %>% group_by(MARRIAGE, DEFAULT_PAYMENT_NEXT_MONTH) %>% count(DEFAULT_PAYMENT_NEXT_MONTH)
marriage #this shows the counts of people per group.

barchart(marriage$n~marriage$DEFAULT_PAYMENT_NEXT_MONTH, ylab='Number of people', xlab='Payment default status', main='Influence of marital status on default payments', groups=marriage$MARRIAGE, auto.key=list(space='right', title='Marriage status', cex.title=0.8))

#----------------------------------------------

#View education status of those who did/didn not default.
education <-credit %>% group_by(EDUCATION, DEFAULT_PAYMENT_NEXT_MONTH) %>% count(DEFAULT_PAYMENT_NEXT_MONTH)
education #this shows the counts of people per group.

barchart(education$n~education$DEFAULT_PAYMENT_NEXT_MONTH, ylab='Number of people', xlab='Payment default status', main='Influence of educational status on default payments', groups=education$EDUCATION, auto.key=list(space='right', title='Educational status', cex.title=0.8))


#What about the link between total bill amount and default status, or total bill amount and how many months they pushed back the payment?




```
1. Are younger people more or less likely to default on payments?
```{r}
#Check normalcy of the data
normalityTest(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, test="ad.test", data=credit)
#The data are normally distributed (p<<0.75)
#Check heteroscedasticity
with(credit, tapply(AGE, DEFAULT_PAYMENT_NEXT_MONTH, var, na.rm=TRUE))
bartlett.test(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, data=credit)
#Data are homoscedastic; p<<0.05


#ANOVA to determine difference of mean age in each group (default vs not default)
AnovaModel.1 <- aov(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, data=credit)
summary(AnovaModel.1)
with(credit, numSummary(AGE, groups=DEFAULT_PAYMENT_NEXT_MONTH, 
  statistics=c("mean", "sd")))

```
```{r}
#Tukey test for age difference
age_signif<-TukeyHSD(AnovaModel.1)
age_signif
```
The above statistical analysis shows us that there is a significant difference in mean age between 'default' and 'will not default' groups: a small difference of 0.3084 years (about 3.7 months), with the slightly younger group tending to default on their payments more often than the slightly older group. 


2. Are higher-educated people more or less likely to default?
```{r}
#Performing pearson Chi-sq test of independence - 
#Actually no, performing the fisher's exact test, since it works better for smaller 'unbalanced' data such as this where some columns have 0 entries.

#First tabulating the contingency table of the two variables
educ_conting=table(credit$EDUCATION, credit$DEFAULT_PAYMENT_NEXT_MONTH)
educ_conting #This gives the counts of each category in each 'axis' specified
#Null hypothesis: "education level is independent of whether or not someone will default, at the 0.05 significance level."

fisher.test(educ_conting, simulate.p.value=TRUE)

#pairwise test
pairwiseNominalIndependence(educ_conting)
```
The results of the Fisher's Exact test show that there is a highly significant interaction between education level and probability that someone will default on their credit card bill (p<<<0.05). 

The pairwise post-hoc test shows significant differences in likelihood of defaulting on credit payments between nearly every pair of education levels. (NEED TO DOUBLE CHECK THIS? WHICH STATISTIC P VALUE SHOULD WE BE LOOKING AT?)


3. How is credit limit associated with likelihood to default? How is credit limit associated with age and education level?
```{r}

```

4. At what bill value is someone more likely to default on their payment? Are there any demographic trends that emerge there?
```{r}

```
5. Is the amount of one single bill more influential than whether or not the  person delayed their last bill in deciding whether someone will default?
```{r}

```

6. How does marital status affect probability of defaulting?
```{r}

```

Visualization of bill amount due vs. payment status
```{r}
plot(credit$PAY_1, credit$BILL_AMT1, main='Payment status or deferral by bill amount', xlab='Payment status code', ylab='Bill amount (NT dollar)', col='darkgreen')
```
Visualization of bill amount due vs. amount paid
```{r}
#par(mfrow=c(3,2))
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT1, y=credit$BILL_AMT1))+
  ggtitle('Amount paid vs. Amount Due, Bill 1')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT2, y=credit$BILL_AMT2))+
  ggtitle('Amount paid vs. Amount Due, Bill 2')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT3, y=credit$BILL_AMT3))+
  ggtitle('Amount paid vs. Amount Due, Bill 3')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT4, y=credit$BILL_AMT4))+
  ggtitle('Amount paid vs. Amount Due, Bill 4')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT5, y=credit$BILL_AMT5))+
  ggtitle('Amount paid vs. Amount Due, Bill 5')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT6, y=credit$BILL_AMT6))+
  ggtitle('Amount paid vs. Amount Due, Bill 6')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")


```
Now all on one plot
```{r}
#all on one plot
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT1, y=credit$BILL_AMT1, col='Bill 1'))+
  geom_point(mapping=aes(x=credit$PAY_AMT2, y=credit$BILL_AMT2, col='Bill 2'))+
  geom_point(mapping=aes(x=credit$PAY_AMT3, y=credit$BILL_AMT3, col='Bill 3'))+
  geom_point(mapping=aes(x=credit$PAY_AMT4, y=credit$BILL_AMT4, col='Bill 4'))+
  geom_point(mapping=aes(x=credit$PAY_AMT5, y=credit$BILL_AMT5, col='Bill 5'))+
  geom_point(mapping=aes(x=credit$PAY_AMT6, y=credit$BILL_AMT6, col='Bill 6'))+
  xlab('Amount paid, NT dollars')+
  ylab('Bill amount, NT dollars')+
  ggtitle('Amount paid vs. Amount due, All Bills')+
  labs(color='Bill Number')
```
Is this relationship linear?
```{r}
par(mfrow=c(2,2))
linreg <- lm(PAY_AMT1~BILL_AMT1, data=credit)
plot(linreg)
#No, the relationship is nonlinear. It appears to follow an inverse exponential curve.
```
Is there a pattern in bill repayment over time? It seems as though people tend to start out paying less of the bill at once, then later on in the billing cycle, they start to pay more than their current bill - to make up for last time?

What about analysis of the total amount due during the 6 months prior, and total amount paid? Who is still in-debt by the time bill 7 rolls around?
```{r}
credit_totals<-credit %>% group_by(ID)%>% mutate(totaldue=sum(BILL_AMT1, BILL_AMT2, BILL_AMT3, BILL_AMT4, BILL_AMT5, BILL_AMT6), totalpaid=sum(PAY_AMT1, PAY_AMT2, PAY_AMT3, PAY_AMT4, PAY_AMT5, PAY_AMT6), totaldebt=totaldue-totalpaid)
head(credit_totals)
#This new dataframe shows the total due, total paid, and total debt amount at the end of the 6 month period, per person. 
```
```{r}
summary(credit_totals)
```
The summary tables show that the mean 'total debt' is \$238,210, with a minimum of a negative amount meaning an overpayment (of \$2,671,514) and a maximum debt of \$4,116,080.
```{r}
#Visualize total debt.
credit_totals$DEFAULT_PAYMENT_NEXT_MONTH<-as.factor(credit_totals$DEFAULT_PAYMENT_NEXT_MONTH)

boxplot(credit_totals$totaldue, credit_totals$totalpaid, credit_totals$totaldebt)
#Note - this is not how I want to visualize this...having trouble thinking of a better way
```

```{r}
ggplot(data=credit_totals)+
  geom_point(mapping=aes(x=totalpaid, y=totaldue))+
  ggtitle("6-month total payments vs. total dues")+
  ylab("Total bill amount due (sum of 6 months, NT dollar)")+
  xlab("Total amount paid over 6 months, NT dollar")
```
Based on the results of this summary graph, it looks like by the end of the study, most people were still in debt to the bank as they owed more than they had paid.

Perhaps we can cluster the respondents into 2 groups: One who is 'in debt' and one who is not?
```{r}

```


