---
title: "SQL-NoSQL analysis"
author: "Kathleen DeBrota"
date: "April 24, 2019"
output: pdf_document
---

This document is for some analysis and visualizations of the credit card data set for SQL/NoSQL final project. 

Guiding questions: What kind of person is most likely to default on their credit card payments? What demographic factor(s) are most important in predicting this? 

```{r}
library(readxl)
library(dplyr)
library(lattice)
library(rcompanion)
library(ggplot2)
library(Rcmdr)
library(gplots)
library(corrplot)
library(factoextra)
library(rgl)
library(purrr)
```

```{r}
setwd("C:/Users/kdebr/Documents/GitHub/sql-nosql")
credit<-read.csv(file='cleanData.csv')
head(credit)
```
Exploratory views of the data.
```{r}
summary(credit)
```
More of the respondents were women (18,112 vs 11,888 men). The majority were undergraduates (14,030) with some graduates (10,585) and fewer who were high-school educated, uneducated, unknown, or 'other'. The majority were not married: 15,964, but many were married: 13,659. A few were 'other' or unknown for marital status. The mean age of the respondents was 35.49 years, with the youngest participants at 21 years and oldest at 79 years. 
```{r}
#view thistograms of demographic data
hist(credit$AGE, main='Histogram of age distribution of respondents', ylab='Frequency', xlab='Age in years')

#Boxplot of the payment status for each of the 6 payments.
boxplot(credit$PAY_1, credit$PAY_2, credit$PAY_3, credit$PAY_4, credit$PAY_5, credit$PAY_6, main='Payment status of each of 6 payments', ylab='Status code', xlab='Payment number')

#Boxplot of payment amount for each of 6 payments.
boxplot(credit$BILL_AMT1, credit$BILL_AMT2, credit$BILL_AMT3, credit$BILL_AMT4, credit$BILL_AMT5, credit$BILL_AMT6, main='Bill amount of each of 6 payments', ylab='Bill amount (NT dollars)', xlab='Payment number')



```
## Separate those who defaulted and who did not. Visualize general demographic differences.
```{r}
will_default<-subset(credit, DEFAULT_PAYMENT_NEXT_MONTH==1)
z3countdef<-table(will_default$AGE, will_default$MARRIAGE)

will_pay<-subset(credit, DEFAULT_PAYMENT_NEXT_MONTH==0)
#countpay<- table(will_pay$AGE, will_pay$MARRIAGE)

credit$MARRIAGE<-as.factor(credit$MARRIAGE)
credit$DEFAULT_PAYMENT_NEXT_MONTH <-as.factor(credit$DEFAULT_PAYMENT_NEXT_MONTH)

par(mfrow=c(1,2))
#View ages of those who did and did not default.
hist(will_default$AGE, main='Age distribution: Will default', xlab='Age (years)', ylab='Frequency')
hist(will_pay$AGE, main='Age distribution: Will not default', xlab='Age(years)', ylab='Frequency')

#----------------------------------------------

#View marital status of those who did/didn't default.
par(mfrow=c(1,1))
marriage <-credit %>% group_by(MARRIAGE, DEFAULT_PAYMENT_NEXT_MONTH) %>% count(DEFAULT_PAYMENT_NEXT_MONTH)
marriage #this shows the counts of people per group.

#This is split by payment default status
ggplot(marriage, aes(fill=MARRIAGE, x=DEFAULT_PAYMENT_NEXT_MONTH, y=n))+
  geom_bar(stat='identity', position='fill')+
  ggtitle("Default payment status by marital status")+
  ylab("Fraction of total")+
  xlab('Default payment status')+
  labs(fill='Marital status')

#This is split by marriage status
ggplot(marriage, aes(fill=DEFAULT_PAYMENT_NEXT_MONTH, x=MARRIAGE, y=n))+
  geom_bar(stat='identity', position='fill')+
  ggtitle("Default payment status by marital status")+
  ylab("Fraction of total")+
  xlab('Marital status')+
  labs(fill='Default payment \nstatus')

#barchart(marriage$n~marriage$DEFAULT_PAYMENT_NEXT_MONTH, ylab='Number of people', xlab='Payment default status', main='Influence of marital status on default payments', stacked=TRUE, groups=marriage$MARRIAGE, auto.key=list(space='right', title='Marriage status', cex.title=0.8))

#----------------------------------------------

#View education status of those who did/did not default.
education <-credit %>% group_by(EDUCATION, DEFAULT_PAYMENT_NEXT_MONTH) %>% count(DEFAULT_PAYMENT_NEXT_MONTH)
education #this shows the counts of people per group.

#This is split by education status
ggplot(education, aes(fill=DEFAULT_PAYMENT_NEXT_MONTH, x=EDUCATION, y=n))+
  geom_bar(stat='identity', position='fill')+
  ggtitle("Default payment status by educational status")+
  ylab("Fraction of total")+
  xlab('Educational status')+
  labs(fill='Default payment \nstatus')

#barchart(education$n~education$DEFAULT_PAYMENT_NEXT_MONTH, ylab='Number of people', xlab='Payment default status', main='Influence of educational status on default payments', groups=education$EDUCATION, auto.key=list(space='right', title='Educational status', cex.title=0.8))


#What about the link between total bill amount and default status, or total bill amount and how many months they pushed back the payment?

#----------------------------------------------

#View sex of those who did or did not default

sex <-credit %>% group_by(SEX, DEFAULT_PAYMENT_NEXT_MONTH) %>% count(DEFAULT_PAYMENT_NEXT_MONTH)
sex #this shows the counts of people per group.

#This is split by education status
ggplot(sex, aes(fill=DEFAULT_PAYMENT_NEXT_MONTH, x=SEX, y=n))+
  geom_bar(stat='identity', position='fill')+
  ggtitle("Default payment status by sex")+
  ylab("Fraction of total")+
  xlab('Sex')+
  labs(fill='Default payment \nstatus')

#----------------------------------------------

#View credit limit of defaulters and not defaulters

bwplot(credit$LIMIT_BAL~credit$DEFAULT_PAYMENT_NEXT_MONTH, xlab='Default payment status', ylab='Credit limit, NT dollar', main='Credit limit by default payment status')

```
## Are younger people more or less likely to default on payments?
```{r}
#Check normalcy of the data
normalityTest(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, test="ad.test", data=credit)
#The data are NOT normally distributed (p<<0.05)
#Check heteroscedasticity
with(credit, tapply(AGE, DEFAULT_PAYMENT_NEXT_MONTH, var, na.rm=TRUE))
bartlett.test(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, data=credit)
#Data are homoscedastic; p<<0.05

#Run a non parametric test to check differences of means
with(credit, tapply(AGE, DEFAULT_PAYMENT_NEXT_MONTH, median, na.rm=TRUE))
kruskal.test(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, data=credit)

# #ANOVA to determine difference of mean age in each group (default vs not default)
# AnovaModel.1 <- aov(AGE ~ DEFAULT_PAYMENT_NEXT_MONTH, data=credit)
# summary(AnovaModel.1)
# with(credit, numSummary(AGE, groups=DEFAULT_PAYMENT_NEXT_MONTH, 
#   statistics=c("mean", "sd")))


```
the Kruskal-Wallis test shows there is NO significant difference in the mean ages between groups (did and did not default). 


## Are higher-educated people more or less likely to default?
```{r}
#Performing pearson Chi-sq test of independence.
#Null hypothesis: "education level is independent of whether or not someone will default, at the 0.05 significance level."

#First tabulating the contingency table of the two variables
educ_conting=table(credit$EDUCATION, credit$DEFAULT_PAYMENT_NEXT_MONTH)
educ_conting 
balloonplot(t(educ_conting), main='Education level frequency by default payment status', xlab='Default payment status', ylab='Educational status', label=FALSE, show.margins=FALSE)
#This gives the counts of each category in each 'axis' specified

#Chi-squared test
chi_sq_test<-chisq.test(educ_conting)
chi_sq_test
#round(chi_sq_test$expected,2)

#To determine the nature of the dependence between rows (education) and columns (payment status), use chi-sq residuals (Pearson residuals)
chisq_resid<-round(chi_sq_test$residuals,3)
chisq_resid

#Visualization of the residuals shows the positive (blue) and negative (red) correlations between the row and column variables.
corrplot(chisq_resid, is.cor=FALSE)
```
The results of the chi square test show that there is a highly significant interaction between education level and probability that someone will default on their credit card bill (p<<<0.05). 

We can also see from the chi-squared test and analysis of the residuals in the corrplot that there is a strong positive association between those who defult and those with either a high school or undergraduate education. Meanwhile, defaulting is negatively associated with having a graduate education or 'uneducated, other, or unknown' education level.

## How does marital status affect probability of defaulting?
```{r}
#Performing pearson Chi-sq test of independence.
#Null hypothesis: "marital status is independent of whether or not someone will default, at the 0.05 significance level."

#First tabulating the contingency table of the two variables
mar_conting=table(credit$MARRIAGE, credit$DEFAULT_PAYMENT_NEXT_MONTH)
mar_conting 
balloonplot(t(mar_conting), main='Marital status frequency by default payment status', xlab='Default payment status', ylab='Marital status', label=FALSE, show.margins=FALSE)
#This gives the counts of each category in each 'axis' specified

#Chi-squared test
chi_sq_test<-chisq.test(mar_conting)
chi_sq_test
#round(chi_sq_test$expected,2)

#To determine the nature of the dependence between rows (education) and columns (payment status), use chi-sq residuals (Pearson residuals)
chisq_resid<-round(chi_sq_test$residuals,3)
chisq_resid

#Visualization of the residuals shows the positive (blue) and negative (red) correlations between the row and column variables.
corrplot(chisq_resid, is.cor=FALSE)
```
## Is default status determined by sex?
```{r}
#Performing pearson Chi-sq test of independence.
#Null hypothesis: "sex is independent of whether or not someone will default, at the 0.05 significance level."

#First tabulating the contingency table of the two variables
sex_conting=table(credit$SEX, credit$DEFAULT_PAYMENT_NEXT_MONTH)
sex_conting 
balloonplot(t(sex_conting), main='Marital status frequency by default payment status', xlab='Default payment status', ylab='Sex', label=FALSE, show.margins=FALSE)
#This gives the counts of each category in each 'axis' specified

#Chi-squared test
chi_sq_test<-chisq.test(sex_conting)
chi_sq_test
#round(chi_sq_test$expected,2)

#To determine the nature of the dependence between rows (education) and columns (payment status), use chi-sq residuals (Pearson residuals)
chisq_resid<-round(chi_sq_test$residuals,3)
chisq_resid

#Visualization of the residuals shows the positive (blue) and negative (red) correlations between the row and column variables.
corrplot(chisq_resid, is.cor=FALSE)
```

## Visualization of bill amount due vs. payment status
```{r}
plot(credit$PAY_1, credit$BILL_AMT1, main='Payment status or deferral by bill amount', xlab='Payment status code', ylab='Bill amount (NT dollar)', col='darkgreen')
```
## Visualization of bill amount due vs. amount paid
```{r}
#par(mfrow=c(3,2))
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT1, y=credit$BILL_AMT1))+
  ggtitle('Amount paid vs. Amount Due, Bill 1')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT2, y=credit$BILL_AMT2))+
  ggtitle('Amount paid vs. Amount Due, Bill 2')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT3, y=credit$BILL_AMT3))+
  ggtitle('Amount paid vs. Amount Due, Bill 3')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT4, y=credit$BILL_AMT4))+
  ggtitle('Amount paid vs. Amount Due, Bill 4')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT5, y=credit$BILL_AMT5))+
  ggtitle('Amount paid vs. Amount Due, Bill 5')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT6, y=credit$BILL_AMT6))+
  ggtitle('Amount paid vs. Amount Due, Bill 6')+
  ylab("Bill amount, NT dollars")+
  xlab("Amount paid, NT dollars")


```
Now all on one plot
```{r}
#all on one plot
ggplot(data=credit)+
  geom_point(mapping=aes(x=credit$PAY_AMT1, y=credit$BILL_AMT1, col='Bill 1'))+
  geom_point(mapping=aes(x=credit$PAY_AMT2, y=credit$BILL_AMT2, col='Bill 2'))+
  geom_point(mapping=aes(x=credit$PAY_AMT3, y=credit$BILL_AMT3, col='Bill 3'))+
  geom_point(mapping=aes(x=credit$PAY_AMT4, y=credit$BILL_AMT4, col='Bill 4'))+
  geom_point(mapping=aes(x=credit$PAY_AMT5, y=credit$BILL_AMT5, col='Bill 5'))+
  geom_point(mapping=aes(x=credit$PAY_AMT6, y=credit$BILL_AMT6, col='Bill 6'))+
  xlab('Amount paid, NT dollars')+
  ylab('Bill amount, NT dollars')+
  ggtitle('Amount paid vs. Amount due, All Bills')+
  labs(color='Bill Number')
```
Is this relationship linear?
```{r}
par(mfrow=c(2,2))
linreg <- lm(PAY_AMT1~BILL_AMT1, data=credit)
plot(linreg)
#No, the relationship is nonlinear. It appears to follow an inverse exponential curve.
```
Is there a pattern in bill repayment over time? It seems as though people tend to start out paying less of the bill at once, then later on in the billing cycle, they start to pay more than their current bill - to make up for last time?

## Total amounts due and total amounts paid analysis
What about analysis of the total amount due during the 6 months prior, and total amount paid? Who is still in-debt by the time bill 7 rolls around?
```{r}
credit_totals<-credit %>% group_by(ID)%>% mutate(totaldue=sum(BILL_AMT1, BILL_AMT2, BILL_AMT3, BILL_AMT4, BILL_AMT5, BILL_AMT6), totalpaid=sum(PAY_AMT1, PAY_AMT2, PAY_AMT3, PAY_AMT4, PAY_AMT5, PAY_AMT6), totaldebt=totaldue-totalpaid)
head(credit_totals)
#This new dataframe shows the total due, total paid, and total debt amount at the end of the 6 month period, per person. 
```
```{r}
summary(credit_totals)
```
The summary tables show that the mean 'total debt' is \$238,210, with a minimum of a negative amount meaning an overpayment (of \$2,671,514) and a maximum debt of \$4,116,080. - note these are NT dollars...not US dollars.
```{r}
#Visualize total debt.
credit_totals$DEFAULT_PAYMENT_NEXT_MONTH<-as.factor(credit_totals$DEFAULT_PAYMENT_NEXT_MONTH)

#barchart(credit_totals$totaldue, credit_totals$totalpaid, credit_totals$totaldebt)
#Note - this is not how I want to visualize this...having trouble thinking of a better way
```

```{r}
ggplot(data=credit_totals)+
  geom_point(mapping=aes(x=totalpaid, y=totaldue, col=DEFAULT_PAYMENT_NEXT_MONTH))+
  ggtitle("6-month total payments vs. total dues")+
  ylab("Total bill amount due (sum of 6 months, NT dollar)")+
  xlab("Total amount paid over 6 months, NT dollar")+
  labs(color='Default payment \nstatus')
 
```


## Clustering
Can we cluster the above graph into two types of people - those who are in debt vs. those who are not?
```{r}
#K-means clustering

#first subset only the numeric columns of the data.
numeric<-cbind(unlist(lapply(credit, is.numeric)))
numeric<-credit[,numeric]
#head(numeric)

#Generate the elbow plot to determine best number of clusters
model <- kmeans(x=numeric, centers=2)
model$tot.withinss
tot_withinss<- map_dbl(1:20, function(k){
  model <- kmeans(x=numeric, centers=k)
  model$tot.withinss
})

elbow_df <- data.frame(
  k=1:20,
  tot_withinss=tot_withinss
)
```
now plot the elbow plot
```{r}
ggplot(elbow_df, aes(x=k, y=tot_withinss))+
  geom_point() +
  geom_line()+
  scale_x_continuous(breaks=1:200) +
  xlim(0,15) +
  ggtitle("Elbow plot to determine optimim k value")+
  xlab("Number of clusters, k")+
  ylab("Total within-cluster-sum-of-squares")
```
Based on this plot it looks like 4 clusters is probably ideal.
Run the k-means clustering with 4 clusters:
```{r}
#Running the cluster analysis with 3 clusters.
km <- kmeans(numeric, centers=3, iter.max=50)

#Visualizing the clusters.
fviz_cluster(km, numeric, geom=c("point"), show.clust.cent = TRUE)
```
This shows the clusters outlined as the kmeans algorithm identified them. The 2 principal components explain 31.1% and 19.4% of the variation in the data respectively. 
```{r}
#PCA analysis of these clusters.
pc<-princomp(numeric, cor=TRUE, scores=TRUE)
summary(pc)
plot(pc, type='lines')
biplot(pc)
```
From these plots of the PCA analysis, it looks like there are 3 major components which are making up the majority of the variation in the data. It does appear that all the BILL_AMTs and PAY_#'s (payment statuses) are correlated heavily with each other.
```{r}
#plot3d(pc$scores[,1:3], col=as.integer(credit$DEFAULT_PAYMENT_NEXT_MONTH))
#the above works and makes an awesome graph, but a little hard to interpret since they are components.

plot3d(credit$PAY_1, credit$BILL_AMT1, credit$PAY_AMT1, col=as.integer(credit$DEFAULT_PAYMENT_NEXT_MONTH), xlab='Payment status', ylab='Bill amount', zlab='Payment amount')
```
other visualizations of the cluster
```{r}
#Visualizing the clusters.
fviz_pca_ind(pc, geom='point', col.ind=credit$DEFAULT_PAYMENT_NEXT_MONTH)
```
```{r}
#visualize the variables.
fviz_pca_var(pc)
```
We can see the three major variables contributing to the clustering based on the above variable graph: generally, 'payment status', 'bill amount', and 'payment amount'. 
